Modelling last glacial cycle ice dynamics in the Alps

### 2018-07-23 articles/cycle/response-referee3.tex

    Unlike periodic surges \citep[as described by, e.g.,][]{Sevestre.Benn.2015},
    we interpret the model results as triggered by an initial perturbation of
    mass (and energy) caused by rapid climate cooling and inefficient mass
    transfer to lower elevations. We have rephrase the following sentence:

    \msquote{%
        Several Alpine lobes surge and overshoot their balanced extent before
        thinning and receding towards the mountains as they warm towards
        thermodynamical equilibrium.}

    \referee{\textbf{p.~19, l.~9--12:}
        On trimlines: I think you justify well why you haven't yet modelled the
        sensitivity to your trimline result - previous work backs this up. But,
        I think you should consider the following: Adding a plus/minus to your
        results to reflect the uncertainty.  The importance of resolution -
        many trimlines will be below the resolution of your model, so perhaps
        aren't resolved enough for model-data comparison. They may have acted
        to deflect ice flow around mountain peaks for example. This resolution
        caveat should be mentioned. I would be surprised if all trimlines are
        subglacial transitions as this paper suggests - perhaps you need to
        directly challenge the geochronological community to find better
        trimline constraints (sub/supra) as a statement in this paper. Your
        mean value of 861 m is unrepresentative of your sample. Your sample is
        highly skewed, so a modal value (1050 m ish from Fig 6) is more
        appropriate. A similar finding with a similar approach has already been
        found for the British Isles, with the added constraint of GIA
        observations. I suggest referencing Kuchar et al. 2011 for this
        reason.}

    This is a valid point. However we found an error in our figure: the data
    plotted on the histogram (Fig.~3b) were different from that in the scatter
    plot (Fig.~3a). These data corresponded to raw interpolated ice
    thicknesses, i.e. they missed a correction for altitude shifts between our
    initial basal topography and the observed trimline elevations.

    The corrected histogram is closer to symmetry. A slight skewness remains,
    but this is not really a result of our simulations. In fact the
    distribution of observations itself is skewed, with more trimlines observed
    at higher elevations, and the differences between the modelled ice surface
    and trimlines are slightly higher at higher elevations (not plotted because
    correlation is poor).

    The modal value (now ca.~950\,m) corresponds to a cluster of trimline
    locations around 2850\,m in the Aletsch Glacier (south of Jungfrau) region
    where trimlines are well visible in the landscape. We prefer not to
    use it because the number of trimlines mapped is not necessarily
    proportional to the lenght, but may be instead depend on other factors such
    as bedrock lithology and post-glacial erosion.

    We now use the standard deviation (197\,m) as a measure of variance, but
    not uncertainty, which would require more sensitivity tests. We also
    mention resolution as one of several caveats in the text:

    \msquote{%
        The remaining discrepancies may relate to temporal migrations of the
        basal thermal boundary, an absence of sliding in warm-based areas, and
        levelling of small-scale topographic features in the 1\,km horizontal
        grid.}

    We added a reference to the study by \citet{Kuchar.etal.2012} although this
    approach is not doable for the Alps.

    \msquote{%
        Unfortunately, validation through the bedrock uplift rates
        \citep[cf.][]{Kuchar.etal.2012} is not doable in the Alps due its lower
        values, active tectonics and uncertainties on geological properties
        \citep[cf.][]{Mey.etal.2016}.}

### 2018-07-23 articles/cycle/alps-cycle.tex
Address referee 3 comment on trimlines.

    In the upper Rhone Valley, the maximum ice surface elevation reached
    during MIS~2 in our simulation, corrected for bedrock deformation, is
    consistently modelled to have occured several hundred metres above the
    observed trimline elevations (Fig.~\ref{fig:trimlines}a and b), with a mean
    difference of 861\,m (Fig.~\ref{fig:trimlines}b), wich is significant in
    respect to the modelled regional ice thickness \todo{add a number}. The result is
    somewhat dependent on uncertain basal sliding and ice rheological parameters
    to which the model sensitivity was not tested here. However, a previous
    sensitivity study conducted with PISM and a very similar model set-up on
    the Cordilleran ice sheet show that varying ice rheological parameters
    resulted in relative errors in modelled total ice volume of 25\,\unit{\%}
    with regard to ice rheological parameters and 14\,\unit{\%} regarding basal
    sliding parameters \citep[Fig.~7]{Seguinot.etal.2016}. In addition,
    regional sensitivity tests in the Rhone Valley show that a modelled ice
    surface compatible with the trimline elevations is incompatible with the
    documented overspilling flow to the south across the Simplon pass
    \citep{Becker.etal.2017}.

@@

    In the upper Rhone Valley, observed trimlines are often located near the
    LGM cold-temperate basal thermal transition, or in cold-based areas
    (Fig.~\ref{fig:trimlines}c). The remaining discrepancies may relate to
    temporal migrations of the basal thermal boundary, an absence of sliding in
    warm-based areas, and levelling of small-scale topographic features in the
    1\,km horizontal grid. They call for more detailed comparisons
    spanning the entire Alpine range and specific sensitivity studies to
    relevant basal sliding and ice rheological parameters, and to the uncertain
    subglacial topography. However, the presence of an upper layer of
    cold ice during the LGM, already found at high altitude in the much warmer
    climate of today \citep[e.g.,][]{Suter.etal.2001, Bohleber.etal.2017}, is
    inevitable \citep[Fig.~3]{Haeberli.Schluchter.1987}.
    In this context, our results challenge the assumption that
    Alpine trimlines mark the maximum upper ice surface elevation of the LGM
    ice cover and call for a more accurate estimation of the thickness of the
    upper layer of cold ice in the Alps.

>>>>

    In the upper Rhone Valley, the maximum ice surface elevation reached
    during MIS~2 in our simulation, corrected for bedrock deformation, is
    consistently modelled to have occured several hundred metres above the
    observed trimline elevations (Fig.~\ref{fig:trimlines}a and b), with a mean
    difference of $861\pm197$\,m (one standard deviation,
    Fig.~\ref{fig:trimlines}b), wich is significant in respect to the modelled
    regional ice thickness \todo{add a number}. The result is somewhat
    dependent on uncertain basal sliding and ice rheological parameters
    to which the model sensitivity was not tested here. However, a previous
    sensitivity study conducted with PISM and a very similar model set-up on
    the Cordilleran ice sheet show that varying ice rheological parameters
    resulted in relative errors in modelled total ice volume of 25\,\unit{\%}
    with regard to ice rheological parameters and 14\,\unit{\%} regarding basal
    sliding parameters \citep[Fig.~7]{Seguinot.etal.2016}. In addition,
    regional sensitivity tests in the Rhone Valley show that a modelled ice
    surface compatible with the trimline elevations is incompatible with the
    documented overspilling flow to the south across the Simplon pass
    \citep{Becker.etal.2017}.

@@

    In the upper Rhone Valley, observed trimlines are often located near the
    LGM cold-temperate basal thermal transition, or in cold-based areas
    (Fig.~\ref{fig:trimlines}c). The remaining discrepancies may relate to
    temporal migrations of the basal thermal boundary, an absence of sliding in
    warm-based areas, and levelling of small-scale topographic features in the
    1\,km horizontal grid. They call for more detailed comparisons
    spanning the entire Alpine range and specific sensitivity studies to
    relevant basal sliding and ice rheological parameters, and to the uncertain
    subglacial topography. Unfortunately, validation through the bedrock uplift
    rates \citep[cf.][]{Kuchar.etal.2012} is not doable in the Alps due its
    lower values, active tectonics and uncertainties on geological properties
    \citep[cf.][]{Mey.etal.2016}. However, the presence of an upper layer of
    cold ice during the LGM, already found at high altitude in the much warmer
    climate of today \citep[e.g.,][]{Suter.etal.2001, Bohleber.etal.2017}, is
    inevitable \citep[Fig.~3]{Haeberli.Schluchter.1987}.
    In this context, our results challenge the assumption that
    Alpine trimlines mark the maximum upper ice surface elevation of the LGM
    ice cover and call for a more accurate estimation of the thickness of the
    upper layer of cold ice in the Alps.
